---
title: "To Dump... or not to dump: comparing strategies for investing lump sums into the S&P"
knit: (function(input_file, encoding) {
  out_dir <- 'docs';
  rmarkdown::render(input_file,
 encoding=encoding,
 output_file=file.path(dirname(input_file), out_dir, 'index.html'))})
author: "Brennan T. Beal"
date: "3/6/2022"
params:
  time_horizon: 60 # 5 year time horizon
  dca_period: 12 # one-year dca
  initial_investment: 40000 # $40k bonus
output:   
  html_document:
    code_folding: hide
    theme: cosmo
    highlight: tango
    toc: true
    toc_float: true
---





```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE)

library(kableExtra)
library(tidyverse)
library(tidyquant)
library(highcharter)

options(dplyr.summarise.inform = FALSE)
# Import data
price_data <- tq_get(
  '^GSPC', 
  get = "stock.prices", 
  # Note, the earliest price from the S&P is 1927
  from = " 1957-01-01")
# Analysis:
# For each month, 
# --- calculate the return of two strategies:
# 1. Dump 30k in the market that day
# 2. Dollar-cost average for 12 months
# Calculate returns after 5 years from the initial date.
# We need to do some work to the imported data:
monthly_price_data <- price_data %>% 
  # Filter for the first day of every month
  filter(day(date) == 1) %>%
  # Handle missing months (this is not ideal... )
  # Take care of some of it by using average of bookend months
  # ... this doesn't catch all and I'm too lazy rn to implement a more elegant solution
  # for the rest, we'll just fill with previous closing value
  # Note, I'm using close because opening pricing data has many more  missing months
  complete(date = seq.Date(min(date), max(date), by="month")) %>%
  mutate(new_close = if_else(is.na(close),
                             slider::slide_dbl(close, mean, .before = 2, .after = -1),
                             close)) %>% 
  fill(new_close) %>% 
  rename(market_date = date)
```
  
```{r define_functions}
# Calculate returns given two dates and the investment
# -- this will be used to map over all dates and some time horizone
calculate_returns <- function(.min_date, .max_date, .investment, .data){
  
  
  # Ideally here, we would use the opening price; however, there are large
  # ... chunks of missing data for opening prices
  initial_close <- .data$new_close[.data$market_date == .min_date]
  final_close <- .data$new_close[.data$market_date == .max_date]
  
  # The return is simply the investment multiplied by the percentage or initial basis
  roi <- .investment*(1-((initial_close-final_close)/initial_close))
  
  return(roi)
  
}
dollar_cost_average <- function(.data, .start_date, .time_period, .horizon, .total_investment){
  
  df       <- .data
  
  # Maximum date to calculate returns should be the complete horizon
  # (ie, if you want to average for a year, but the horizon is 5 years,
  #  ... the return should be for the five year mark)
  max_date <- ymd(.start_date) + months(x = .horizon)
  
  
  # The max investment date is the last date of investing when dollar cost averaging
  # (ie, if one were to invest over 12 moths, the 12th month would be the max investment date)
  # Note, we have to add the time period - 1 because the first month is accounted for
  max_investment_date <- ymd(.start_date) + months(x = (.time_period - 1))
  
  
  # The monthly investment should be the total capital divided by the time period
  # ... for averaging
  monthly_investment <- .total_investment/.time_period
  
  # get the appropriate date vector to map rllling returns over
  date_vec <- df %>% 
    filter(market_date >= lubridate::ymd(.start_date),
           market_date <= max_investment_date) %>% 
    pull(market_date)
  
  monthly_values <- date_vec %>%
    setNames(., nm = as.character(1:(.time_period))) %>% 
    purrr::map_df(.,
               .f = ~calculate_returns(.data = df,
                                       .min_date = ymd(.x),
                                       .max_date = max_date,
                                       .investment = monthly_investment) %>% 
                 as.tibble(),
               .id = "month")
  
  sum(monthly_values$value)
  
}
```

### Tl;dr  
  
***Historically, investing all of one's money at the time of bonus is a stronger strategy; however, this is likely due to the fact that since the beginning of the S&P, the market has mostly done well. In bearish markets, we can see that a dollar-cost averaging approach can sometimes save one a lot of money!***

### Intro  
  
It's that time of year again - bonus time. Assuming all has went well, this is a pretty great time of the year. But every year around this time, I find myself asking "what should I do with this?" My inclination is that it really doesn't matter... save it... invest it... don't blow it. The usual. Fine, I'll skip the vegas trip this year. But if I'm going to invest my money, how should I do it? There are two pretty intuitive approaches (for a non-finance guy). First, I could just invest it all at once, when it's distributed. But then again, if the market is bad, maybe I should dollar-cost average (DCA) over a period of time so as not to catch a falling knife, as the cool kids say. The purpose of this blog is to explore the nuances of both approaches.  
  
In the following, I'll gather the s&p 500 data from 1957. Then, I'll use that data to compare each approach (upfront vs DCA) to see what works best. Though, before I get into it, I'll note some upfront limitations. Most notably, I'm not a financial adviser. Second, apparently the opening and closing values for the S&P 500 are not always listed. For values that aren't listed but that have bordering values, I take the average, for those with larger gaps than just one value, I've carried the last observation forward (LOCF). Admittedly, this is not the best approach but i'm trying to answer a question on a lazy Sunday so give me a break.  
  
Now that we know the deal, let's get into it.  
  
### The basics

```{r, warning=FALSE}
investment <- params$initial_investment
horizon    <- params$time_horizon # 5 year time horizon
dca_period <- params$dca_period # dollar cost average for a year with lump sum
full_frame <- monthly_price_data %>% 
  # Max date we can map to is the maximum date in the set minus our time horizon
  filter(market_date <= max(market_date) - months(x = horizon)) %>% 
  mutate(
    
    # Employ our upfront approach
    upfront_approach = map_dbl(market_date,
                               ~calculate_returns(
                                 .data = monthly_price_data,
                                 .min_date = .x,
                                 .max_date = ymd(.x) + months(horizon),
                                 .investment = investment)),
    
    # Employ our Dollar-Cost Averaging Approach
    dca_approach     = map_dbl(market_date,
                               ~dollar_cost_average(
                                 .data = monthly_price_data,
                                 .start_date = .x,
                                 .time_period = dca_period,
                                 .horizon = horizon,
                                 .total_investment = investment
                               )),
    
    # Find the difference between the two
    diff = upfront_approach - dca_approach,
    
    # This flag will be used to summarise the proportion of times the upfront strategy is "correct"
    upfront_correct = if_else(upfront_approach >= dca_approach, 1, 0),
    
    # Text of favored strat
    favored = if_else(upfront_correct == 1, "Upfront", "DCA")
    
  ) 
comparison_frame <- full_frame %>% 
  mutate(favored = if_else(upfront_correct == 1, "Upfront", "DCA")) %>% 
  select(market_date, close, upfront_approach, dca_approach, diff) %>% 
  as.data.frame()

xts_frame <- xts(comparison_frame[-1], order.by = comparison_frame$market_date)
```

The first thing I want to do is check out the S&P data as a whole. I think this can be a good reference point for numbers that come up. So, here's the S&P over time...    
  
```{r}
stock_data <- price_data %>% 
  select(date, close)
stock_data_ts <- xts(stock_data[-1], order.by = stock_data$date)
highchart(type = "stock") %>% 
   hc_add_series(data = stock_data_ts$close) %>% 
  highcharter::hc_xAxis(
    labels = list(format = "{value: %Y}"))  %>% 
  hc_tooltip(
    pointFormat = 'Date: {point.x: %Y-%m-%d}<br>
                            Value: ${point.y:,.2f}',
    headerFormat = '')  %>%
  hc_title(
    text = "S&P Closing Price Over Time",
    align = "left"
  ) 
```

There's nothing surprising here. The dot com bubble and 2008 stick out. So let's define some strategies and try them out in our safe space: the past.  
  
#### Strategy 1  
  
Strategy 1 will be an upfront approach, and will be referred to as that going forward. What this means is that we'll take our bonus (or any lump sum) and simply dump it in the market as soon as we get it. For simplicity, I'm assuming that we're putting our money in at the first of each month. I'll take the lump sum, and simulate investing it all at once at the beginning of each month, and we'll see where we are at the end of five years (the five-year returns).  
  
#### Strategy 2  
  
For strategy 2, we'll take a lump sum and divide it by 12. We'll then take that quotient and invest it each subsequent month for 12 months (dollar-cost average) and then let it sit for the remaining four years (for a five-year return).
  
### Analysis  
  
For the analysis I want to know a couple key questions:  
  
    1. How often is the upfront approach "the best"
    2. When the two approaches differ, how much do they differ by?
    3. When are the two approaches the most different and can we learn something from that?  
  
Let's look at some initial summary statistics to answer the first couple questions:  
  
```{r}
upfront_correct <- full_frame %>% 
  summarise(val = mean(upfront_correct)) %>% 
  pull(val)


kable_prop_making_money <- full_frame %>% 
  # create columns to compare signs after storing data in 'sign_check' below 
  mutate(upfront_gains = if_else(upfront_approach > investment, 1, 0),
         dca_gains     = if_else(dca_approach     > investment, 1, 0)) %>% 
  # store data step outside of the pipe
  {{sign_check <<- .}} %>% 
  pivot_longer(cols = contains("gains"),
               names_to = "Strategy") %>% 
  group_by(Strategy) %>% 
  summarise(`Proportion Exceeding Investment` = scales::percent(mean(value))) %>% 
  mutate(Strategy = c("DCA", "Upfront")) %>% 
  kableExtra::kable(
    caption = "The Percentage of Times a Given Strategy Makes Money"
  ) %>% 
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"))


sign_diff <- sign_check %>% 
  mutate(sign_diff = if_else(upfront_gains != dca_gains, 1, 0)) %>% 
  summarise(sign_diff = mean(sign_diff)) %>% 
  pull(sign_diff)
```
  

  
```{r}
kable_prop_making_money
```

  
```{r}
bind_rows(
         formattable::currency(summary(full_frame$upfront_approach)),
         formattable::currency(summary(full_frame$dca_approach)),
         formattable::currency(summary(full_frame$diff))) %>% 
  bind_cols(
    tibble(
      Strategy = c("Upfront", "DCA", "Absolute Difference")
    )
  ) %>% 
  select(
    Strategy,
    everything()
  ) %>% 
  kableExtra::kable(.,
                    caption = "Summary Statistics for Each Approach as well as Summary Statistics for the Absolute Difference of Each Approach") %>% 
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"))
```
When comparing these approaches, the upfront approach has outperformed the dollar-cost averaging approach `r scales::percent(upfront_correct)` of the time, but it seems like for the most part, the differences are negligible. The summary stats above tell us that ~75% of the time our differences between strategies will be confined to $6k over five years (or ~15% of our initial investment). And no matter what, we can also see that you're at least gaining money over the 5-year period over 80% of the time with either approach and there are only a small percentage of times where one strategy makes money and the other doesn't (`r scales::percent(sign_diff)`). So, this should take some pressure off; however, the outliers make me nervous. 
  
Now that we've got that out of the way, I want the look at these numbers a different way. First, I want to see the five-year returns, by initiation month, and by strategy. Then, I want to see the differences in the two strategies. The first chart below provides the former with the second chart providing information on the latter. In the second chart, the blue represents initiation months where an upfront approach outperformed and the purple represents the opposite.  

```{r}
hchart(full_frame, 
       "column", 
       hcaes(
         x = market_date, 
         y = upfront_approach
       ), 
       color = "#488FB1", 
       yAxis = 0) %>% 
  hc_yAxis_multiples(
    list(
      title = list(
        text = "Upfront"
      ), 
      top = "0%", 
      height = "40%"),
    list(
      title = list(
        text = "DCA"
        ), 
      top = "50%", 
      height = "40%", 
      opposite = TRUE)  
  ) %>%
  hc_add_series(full_frame, 
                "column", 
                color = "#533E85",
                hcaes(x = market_date, y = dca_approach), 
                yAxis = 1) %>% 
  hc_tooltip(
    pointFormat = 'Date: {point.x: %Y-%m-%d}<br>
                            5-Year Returns: ${point.y:,.2f}',
    headerFormat = '') %>%
  hc_title(
    text = "5-Year Return by Strategy and Beginning Month",
    align = "left"
  ) %>% 
  hc_subtitle(
    text = "assuming a $40k investment",
    align = "left"
  )  %>%  
  hc_legend(enabled = F)
```

```{r comparing_approaches}
highcharter::highchart() %>% 
  highcharter::hc_add_series(
    name = "Upfront Approach", 
    data = xts_frame$diff,
    type = "column", 
    color = "#488FB1",
    negativeColor = "#533E85"
  ) %>% 
  hc_tooltip(
    pointFormat = 'Date: {point.x: %Y-%m-%d}<br>
                            Difference: ${point.y:,.2f}',
    headerFormat = '') %>% 
  highcharter::hc_xAxis(
    labels = list(format = "{value: %Y}")) %>% 
  hc_yAxis(
    labels = list(
      formatter = JS(
        "function(){
        if(this.value >= 0){
        byThousand = this.value/1000 + 'k'
        return(byThousand)
        } else {
        byThousand = '-$' + this.value/1000*-1 + 'k'
        return(byThousand)
        }
        }"
      )
    )
  ) %>%
  hc_title(
    text = "Comparing Upfront versus a Dollar-Cost Averaging Approach with a Lump Sum",
    align = "left"
  ) %>%
  hc_subtitle(
    text = "where positive values favor an upfront approach",
    align = "left"
  ) %>%  
  hc_legend(enabled = F)
  
```
  
This is less clear than i'd like, but not unexpected. We know so far that `r scales::percent(upfront_correct)` of the time the upfront approach has better returns, but that most of the time you'll be positive after five years no matter what. We know that numerically and graphically at this point.  
  
So far, I'd say nothing conclusive.   
  
So finally, I want to know: what were the best five months (in terms of differences) favoring the upfront approach versus the best five months favoring the DCA approach in the last 30 years? Knowing these things, we'll be able to look back at our stock over time and form some hypotheses about what strategies work best in certain market conditions.  
  
  
```{r table_of_top_5}
full_frame %>% 
  mutate(favor_flag = if_else(diff < 0, 1, 0)) %>% 
  filter(year(market_date) > 1990) %>% 
  group_by(favor_flag) %>% 
  mutate(abs_diff = abs(diff)) %>% 
  slice_max(abs_diff, n = 5) %>% 
  mutate(Strategy = if_else(favor_flag == 1, "DCA", "Upfront"),
         Return = if_else(favor_flag == 1, dca_approach, upfront_approach)) %>% 
  ungroup() %>% 
  {{top_dates <<- .}} %>% 
  transmute(`Date` = market_date, Difference = scales::dollar(abs_diff), Return = scales::dollar(Return)) %>% 
  kable(
    caption = "Top 5 Dates Where a Given Strategy Outperformed the Other"
  ) %>% 
  kable_styling() %>% 
  pack_rows("Upfront Strategy", 1, 5, label_row_css = "background-color: #488FB1; color: #fff;") %>%
  pack_rows("DCA Strategy", 6, 10,  label_row_css = "background-color: #533E85; color: #fff;") 
```

  
We can see the best performing months (in terms of strategy differences) and the results are unsurprising with a little bit of context about what was going on during those periods. So, the final task is to overlay those five points on our stock market data to see the trends. Additionally, I'll add the best performing strategy for each month in the tooltip for fun.    
  

```{r labeling_top_5}
dates <- top_dates$market_date
text_for_chart <- tibble(
  date = dates,
  title = paste(top_dates$Strategy),
  abs_diff = top_dates$abs_diff,
  text = paste0('Favored Strategy:', title, '<br> Difference ',scales::dollar(abs_diff))
)

highchart() %>% 
   hc_add_series(data = full_frame, id = "stonks",
                 type = 'line',
                 hcaes(x = market_date, y = new_close),
                 tooltip = list(
                   pointFormat = 'Date: {point.market_date: %Y-%m-%d}<br>
                            Value: ${point.new_close:,.2f} <br> Favored Strategy: {point.favored}',
                   headerFormat = ''
                 )) %>% 
  highcharter::hc_xAxis(
    labels = list(format = "{value: %Y}")) %>% 
  hc_title(
    text = "Top 5 Months for each Strategy",
    align = "left"
  ) %>% 
  hc_subtitle(
    text = "in terms of their difference from the other strategy",
    align = "left"
  ) %>% 
  hc_add_series(
    text_for_chart[text_for_chart$title == "DCA",], 
    hcaes(x = date),
    type = "flags", 
    shape = "circlepin",
    onSeries = "stonks",
    color = "#533E85",
    tooltip = list(
                   headerFormat = ''
                 )
  ) %>% 
  hc_add_series(
    text_for_chart[text_for_chart$title == "Upfront",], 
    hcaes(x = date),
    type = "flags", 
    shape = "circlepin",
    onSeries = "stonks",
    color = "#488FB1",
    tooltip = list(
                   headerFormat = ''
                 )
  )  %>% 
  hc_yAxis(
    labels = list(
      formatter = JS(
        "function(){
        if(this.value >= 0){
        byThousand = this.value/1000 + 'k'
        return(byThousand)
        } else {
        byThousand = '-$' + this.value/1000*-1 + 'k'
        return(byThousand)
        }
        }"
      )
    )
  ) %>%
  hc_legend(enabled = FALSE)
```
  
  
Unsurprisingly, it seems like the biggest differences in the two strategies are realized in bullish or bearish markets, with bullish markets favoring an upfront approach. Timing is everything, as they say.  
  
  